---
title: 'ESM 244 Lab 6 Part 1: Spatial kriging'
author: "Allison Horst"
date: "2/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(tmap)
library(gstat)
library(stars)
```

## Read in the California depth to groundwater data

**Data source:** California Department of Water Resources. (2016). Groundwater Depth, California, Fall 2013. California Department of Water Resources. Available at: http://purl.stanford.edu/py212tb3061.

These data are measurements for depth to groundwater recorded in 2013. We'll read in the layer with `read_sf()`, check the CRS, then make a quick plot to start

```{r}
# Read in the depth to groundwater layer: 
ca_dgw <- read_sf(here("ca_dgw"), layer = "F2013_DBGS_Points_20150720_093252") %>% 
  clean_names()

# Check the projection:
st_crs(ca_dgw) # WGS 84 

# Read in the CA county data (TIGER shapefile):
ca_counties <- read_sf(here("ca_counties"), layer = "CA_Counties_TIGER2016") %>% 
  clean_names() %>% 
  select(name)

# Check the projection
st_crs(ca_counties)
ca_counties <- st_transform(ca_counties, st_crs(ca_dgw))

# Make a quick ggplot:
ggplot() +
  geom_sf(data = ca_counties) +
  geom_sf(data = ca_dgw, aes(color = dgbs))
```

## Map these interactively to explore further

```{r}
tmap_mode("view")

tm_shape(ca_dgw) +
  tm_dots("dgbs")
```

Let's just look at depth to groundwater (ft) in and around San Joaquin County. 

```{r}
# Just get sj_county
sj_county <- ca_counties %>% 
  filter(name == "San Joaquin")

# Now keep observations for groundwater depth within that county:
sj_depth <- ca_dgw %>% 
  st_intersection(sj_county)

# Make a quick plot of those: 
ggplot() +
  geom_sf(data = sj_county) +
  geom_sf(data = sj_depth, aes(color = dgbs))
```

### Check for duplicate locations:

```{r}
well_duplicates <- sj_depth %>% 
  get_dupes(latitude, longitude)

well_duplicates

# This will cause problems (0 spatial variance). We're just going to get rid of these problematic wells recorded at the same location: 

sj_depth <- sj_depth %>% 
  filter(!local_well %in% well_duplicates$local_well)

# Check again for duplicates: 
sj_depth %>% 
  get_dupes(latitude, longitude) 

# No more duplicates
```

## Make a variogram: 

```{r}
# Create the variogram
sj_dgw_vgm <- variogram(dgbs ~ 1, data = sj_depth)

# Plot it
plot(sj_dgw_vgm) # Looks variogramish! Increasing variance as observations get further apart. Let's model this...

# Now create the variogram model, & see ?vgm for other model options: 
sj_dgw_vgm_fit <- fit.variogram(sj_dgw_vgm, model = vgm(nugget = 20, psill = 3000, model = "Gau", range = 30))

# Plot them together:
plot(sj_dgw_vgm, sj_dgw_vgm_fit) 

# You can check the sum of squares of residuals for each: 
attr(sj_dgw_vgm_fit, 'SSErr') # ...and you could compare to other models...

```

## Spatial kriging

```{r}

# Make a grid over which we'll krige: 
sj_grid <- st_bbox(sj_county) %>%
  st_as_stars(dx = 0.01, dy = 0.01) %>%
  st_set_crs(4326) %>%
  st_crop(sj_county)

sj_grid

# Look at the grid we made: 
plot(sj_grid)

# Perform ordinary kriging: 
sj_dgw_krige <- krige(dgbs ~ 1, sj_depth, sj_grid, model = sj_dgw_vgm_fit)

# Initial plot of kriging outcome: 
plot(sj_dgw_krige)

# Convert it to a spatial data frame
krige_df <- as.data.frame(sj_dgw_krige) %>% 
  st_as_sf(coords = c("x","y")) %>% 
  drop_na(var1.pred)

st_crs(krige_df) <- 4326

# Then we can use ggplot: 
ggplot(data = krige_df) +
  geom_sf(aes(color = var1.pred)) +
  scale_color_gradient(low = "blue", high = "yellow")
```

...& there are ways to convert this to raster data / aggregate if you want...

## End variogram & kriging demo

